---
title: "An Exploratory Analysis of Cancer Incidence and Mortality to Identify High-Risk Communities and Improve Survival "
author: "Ramiro Cadavid, Pri Nonis, Payman Roghani"
date: "September 24, 20168"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(car)
library(fBasics)
library(xtable)
```


## Introduction

In this project our efforts are focused on the analysis of data included in the csv file provided, to primarily understand the potential relationship between different parameters and the incidences of cancer across counties in the US. The main objectives are:
1. To understand factors that predict cancer mortality rate, with the ultimate aim of identifying communities for social interventions.
2. To determine which interventions are likely to have the most impact.


## Cancer Data

```{r}
Cancer <- read.csv('cancer.csv')
```

```{r collapse = TRUE, comment=NA}
summary(Cancer) #summary statistics
```


```{r collapse = TRUE, comment=NA}
str(Cancer, max.level = 1, strict.width = "wrap")
```


```{r collapse = TRUE, comment=NA}
colnames(Cancer)
cat("  \n")
print(paste0('Number of rows: ', nrow(Cancer)))
print(paste0('Number of columns: ', ncol(Cancer)))

```


The cancer.csv file 29 variables (30 columns, including the first one that has the number of observations) and 3047 observations, where each observation (i.e. row) includes data for a county across the US. The variables are mostly numbers and integers, except for 2 that are factors (`binnedInc` and `Geography`). Below, we have explain the variables in detail and provide our assessment of the quality of the data.  

>>> data on smoking and obesity and other cancer risk factors could've been very helpful

**Variables**

* Cancer data:
    + `avgAnnCount`: The average number of new cancer cases per year per county for years 2009-2013
    + `popEst2015`: Estimated population by county 2015

* Economic status:
    + `medIncome`: Median income per county
    + `povertyPercent`: Percent of population below poverty line
    + `binnedInc`: ???

* Population age and gender:
    + `MedianAge`: Median age per county
    + `MedianAgeMale`: Median age among males per county    
    + `MedianAgeFemale`: Median age among females per county  

* Location:
    + `Geography`: County, State

* Marital status:
    + `PercentMarried`: Percentage of married population
    + `PctMarriedHouseholds`: Percentage of married households per county

* Education:
    + `PctNoHS18_24`: Percentage of 18-24 year old population with no high school education
    + `PctHS18_24`: Percentage of 18-24 year old population with high school education
    + `PctSomeCol18_24`: Percentage of 18-24 year old population with some college education
    + `PctBachDeg18_24`: Percentage of 18-24 year old population with bachelor's degree 
    + `PctHS25_Over`: Percentage of population above 24 years old with high school education
    + `PctBachDeg25_Over`: Percentage of population above 24 years old with bachelor's degree 

* Household size:
    + `AvgHouseholdSize`: Average household size per county

* Employment status:
    + `PctEmployed16_Over`: Percentage of population above 15 years old who have jobs
    + `PctUnemployed16_Over`: Percentage of population above 15 years old with no jobs

* Health insurance coverage:
    + `PctPrivateCoverage`: Percentage of the population with private insurance coverage
    + `PctEmpPrivCoverage`: percentage of the population with employer-sponsored insurance coverage
    + `PctPublicCoverage`: Percentage of the population with public insurance coverage

* Race:
    + `PctWhite`: Percentage of white population by county
    + `PctBlack`: Percentage of African-American population by county
    + `PctAsian`: Percentage of Asian population by county
    + `PctOtherRace`: Percentage of other races by county

* Birth and death rates: 
    + `BirthRate`: Birth rate per county
    + `deathRate`: Death rate per county       


**Evaluation of Dataset and Variables**

Based on the outputs from diagnostic and summary statistics functions that we used above and further analysis explained in later sections of this report, below we describe our evaluation of dataset and its variables. Since definitions of most variables were not provided to us, our first step was to ensure understanding of what such variables represent. We also evaluated the data to identify potentially erroneous values and determine what variables are key to our analysis and whether we the dataset has the right variables to help answer the project questions or we would need to create additional variables needed to achieve that goal.  

>>>from the assignment document: Evaluate the data quality. Are there any issues with the data? Explain how you handled these potential issues. Explain whether any data processing or preparation is required for your data set.
>>>create references between bullet points below and analysis done to support our evaluation/assumptions

* Data time frame: While `avgAnnCount` represents the mean for years 2009-2013, the population by county is for 2015 and other variables do not have date stamps. Ideally all variables should have been from the same time period.

* `avgAnnCount`: There is no definition for incidence rate per county for the `avgAnnCount` variable. Since the sum of all values is 1,847,514 and based on cancer.gov data the average number of cases for years 2009-2013 is 1,617,144, we will assume this variable represents the actual count of new cases. Therefore, in our analysis we created a new variable called "..." to represent the incidence rate of cancer per county (number of new cases per 100,000 people).

```{r collapse = TRUE, comment=NA}
#caclulating the total for avgAnnCount to compare with offical reports by cancer.gov
sum(Cancer$avgAnnCount)
```


```{r collapse = TRUE, comment=NA}
#calculating the ofr mean cancer death count for years 2009-2013 based on cancer.gov data, in order to cofirm our assumption for the definition of avgAnnCount variable  
incidence_cancer <- c(1660290, 1529560, 1596670, 1638910, 1660290)
mean(incidence_cancer)

```


* Through our assessment we realized that the number of cancer new cases (`avgAnnCount`) for 6 counties were greater than the those counties population (`popEst2015`). Looking at the 6 observations, we realized that the the new case count for all these 6 counties is exactly the same number (1962.667684). In fact there are a total of Y counties that have exactly the same average number of new cases, which is probably an erroneous value. We decided to replace all of them with NA in our analysis.


```{r collapse = TRUE, comment=NA}
#cheking the number of observatios, where new case count is greater than the population
sum(Cancer$avgAnnCount > Cancer$popEst2015, na.rm = TRUE) 
```


```{r collapse = TRUE, comment=NA}
Cancer$avgAnnCount[Cancer$avgAnnCount == 1962.667684] <- NA #removing the potentially erroneous number
Cancer$incidenceRate <- Cancer$avgAnnCount / Cancer$popEst2015 * 100000 #creating a new variable: new cases per 100,00 people per county
```


* We checked the `Geography` variable to identify potential duplicates. Since the number of unique values in this column is equal to the total number of observations, there can not be any duplicates in this column.

```{r collapse = TRUE, comment=NA}
length(unique(Cancer[["Geography"]]))
```


* The `binnedInc` variable has 10 levels that seem arbitrary. It is not clear why the income bins have been defined this way.


* The maximum `MedianAge` shows a value of 624, which is clearly a wrong number. We actually identified a total of 30 values in this column that are above 100; therefore, we will replace such values with NA in our analysis.

```{r collapse = TRUE, comment=NA}
age_error = subset(Cancer, MedianAge > 100) #checking the number of erroneous values
nrow(age_error)
```

```{r collapse = TRUE, comment=NA}
Cancer$MedianAge[Cancer$MedianAge > 100] = NA #replacing erroneous values with NA
```


* The minimum `AvgHouseholdSize` is 0.0221, which does not make sense, since we don't expect a household size below 1. There are 61 values in this column that are below 1, which we will replace with NA in our analysis.  

```{r collapse = TRUE, comment=NA}
household_error = subset(Cancer, AvgHouseholdSize < 1) #checking the number of erroneous values
nrow(household_error)
```

```{r collapse = TRUE, comment=NA}
Cancer$AvgHouseholdSize[Cancer$AvgHouseholdSize < 1] = NA #replacing erroneous values with NA
```


* The `PctSomeCol18_24` variable has too many NA values (2285 out 3047). We will need to take this into account during our analysis. 


* It is not clear how the birth rate is calculated and what exactly `BirthRate` represents. Often, the birth rate is defined as childbirths per 1,000 people each year, but applying that here would not give us the right number. For example in Los Angeles County with the population of 10,170,292, there were 124,641 live births in 2015, which translates into a birth rate of 12.25 (BR = (b รท p) X 1,000). However, the birth rate in our data shows a value of 4.7, which is probably the ratio of women aged 15-50 years old who gave birth in 2015 (source: http://www.towncharts.com/California/Demographics/Los-Angeles-County-CA-Demographics-data.html). As a result we didn't use this variable in our analysis.

```{r collapse = TRUE, comment=NA}
#checking the BirthRate value for Los Angeles County
Cancer[1000,'BirthRate']
```


```{r collapse = TRUE, comment=NA}
#Calculating LA County birth rate based on official figures. Formula: BR = (b รท p) X 1,000
124641/10170292*1000
```


* Based on our assessment, we believe the `deathRate` should represent the number of deaths due to cancer per 100,000 population per county. We looked at the figure for Kings County, NY (173.6) and the number in our data is closer to cancer death rate (140.3), as opposed to overall death rate (603.1). Based on this assumption, we also calculated the total death in a new column, calling the variable death_count (`deathRate` * `popEst2015`/100000) and total is 525,347, which is close to the figure reported by cancer.gov (589,430), further confirming our assumption regarding `deathRate` is most probably correct.


```{r collapse = TRUE, comment=NA}
#checking the deathRate for Kings County, NY
Cancer[388, 'deathRate']
```

*Kings County, NY statistics:
2015 population: 2,673,000
2015 death rate (per 100,000 population): 603.1
2015 Cancer death rate (per 100,000 population): 140.3
Sources: DATA USA https://datausa.io/, NY State Dpt of Health https://www.health.ny.gov/

```{r collapse = TRUE, comment=NA}
#comparing total death count in our dataset with official stats reported by officials
Cancer$death_count <- Cancer$deathRate * Cancer$popEst2015/100000
sum(Cancer$death_count)
# 2015 cancer mortality reported by Cancer.gov: 589,430
```


* We assume that the values in `PctEmpPrivCoverage` column represent a subset of values The sum of values in `PctPrivateCoverage` column, since the sum of these two variables in some rows is above 100.
* Also, we assume that there is an overlap between people that have public health insurance and those with private health insurance, since the sum of `PctPrivateCoverage` and `PctPublicCoverage` in some rows is above 100. 

```{r collapse = TRUE, comment=NA}
#adding up health insurance coverage variables, to makes sence of such variables
Cancer$Pct_insured <- Cancer$PctPrivateCoverage + Cancer$PctPublicCoverage
Cancer$Pct_PersonalIsure <- Cancer$PctPrivateCoverage + Cancer$PctEmpPrivCoverage
print('Cancer$Pct_insured')
summary(Cancer$Pct_insured)
print('Cancer$Pct_PersonalIsure')
summary(Cancer$Pct_PersonalIsure)
```

As seen in the summary statistics above, the Max for the 2 variables are above 100. 



* other: removal of outliers? check with team



```{r collapse = TRUE, comment=NA}
#adding 2 separate columns for County and State, in order to State-wide analysis of the data
Cancer <- Cancer %>% separate(Geography, c("County", "State"), sep = ",", remove = FALSE)
```


# Multiavriate analysis

```{r collapse = TRUE, comment=NA}
scatterplotMatrix( ~ incidenceRate + deathRate +
                   + PctPrivateCoverage + PctEmpPrivCoverage + PctPublicCoverage
                   ,diagonal=list(method="histogram"), 
                   data = Cancer, main = "Scatterplot Matrix to Understand the Impact of Insrance Coverage")
```

```{r collapse = TRUE, comment=NA}
cor(Cancer[ , c("incidenceRate", "deathRate",
                 "PctPrivateCoverage", "PctEmpPrivCoverage", "PctPublicCoverage")], 
                  use = "complete.obs")
```

Payman's note:
* There is a positive correlation between incidenceRate and PctPublicCoverage (0.49), while the correlation between incidenceRate and PctPrivateCoverage is almost zero (-0.22 for PctEmpPrivCoverage)
* There is a positive correlation between deathRate and PctPublicCoverage (0.40), while the correlation between incidenceRate and PctPrivateCoverage is negative (-0.36)
* Based on this we can make a conclusion that public health insurance probably results in higher incidence of cancer and mortality
* Caveat: the type of health insurance coverage (public vs private) is often affected by other factors. For example for geographic locations with low average employment/income, we can expect higher public insurance coverage. 
* Note: for future recommendations, we should also consider the major changes in public health insurance coverage due to Affordable Care Act, which aims to increase the quality of care through establishment of pay-for-performance and value-based healthcare policy. 






